---
title: "Getting started with networkScore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with networkScore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`networkScore` computes a permutation-test significance score for a
condition's PCSF network -- either kinase-only, or paired with sensitivity
data ("golden score"). Every network build goes through `networkGen`
(see its own vignette, `vignette("networkGen")`, for the underlying
`prepare_run_params()`/parameter-tracking mechanism this package builds on,
and its "unified grid interface" section for `build_network_grid()`, which
`make_golden_score_kinase()`/`make_golden_score_full()` use internally too --
`spec_cutoff`/`perc_cutoff` are gridded exactly the same way here as they
are for plain generation).

```{r setup}
library(networkGen)
library(networkScore)
```

## A tiny real example

Raw UKA (kinase-activity) data, Tercen-style dotted column names, two
conditions, 8 real genes sampled from the bundled `ppi_networkv12_502_kins`
(the smallest of the three bundled reference networks -- 502 kinases,
~107K edges, vs. 1.1M for the full `ppi_networkv12` -- good enough for a
kinase-activity network and much faster for an example):

```{r toy-uka}
set.seed(1)
real_genes <- sample(unique(c(ppi_networkv12_502_kins$head, ppi_networkv12_502_kins$tail)), 8)
real_genes

toy_uka <- data.frame(
  `x.Sample` = rep(c("DrugA_vs_DMSO", "DrugB_vs_DMSO"), each = 8),
  `x.Kinase Name` = rep(real_genes, 2),
  `x.Median Kinase Statistic` = c(
    c(2.5, -2.0, 1.8, -1.5, 1.2, -1.0, 0.8, -0.5),
    c(-1.8, 2.2, -1.0, 1.5, -0.8, 1.0, -0.5, 0.6)
  ),
  `x.Mean Specificity Score` = 2.0,
  check.names = FALSE
)
head(toy_uka)
```

## Kinase-only score

`nperms_network` is kept small here (2) for a fast-running example; use 50+
for a real analysis. `b = 50` clears PCSF's terminal-inclusion threshold for
this data's percentile-rank-scale prizes (see `vignette("networkGen")`'s
"Key parameters" section for the `b`/`w` economics). `perc_cutoff` is a
vector -- both values are scored, gridded against every condition:

```{r kinase-score}
respath_kinase <- tempfile("kinase_score_")
dir.create(respath_kinase)

result <- make_golden_score(
  uka = toy_uka,
  spec_cutoff = 1.0,
  perc_cutoff = c(0.5, 0.7),
  ppi_network = ppi_networkv12_502_kins,
  b = 50,
  respath = respath_kinase,
  nperms_network = 2
)

result$results
```

## Golden score (paired kinase + sensitivity)

The paired path needs a `contrast` column identifying treatment vs. control
(`clean_uka_to_kinograte_full()` handles both "X vs control" and "control vs
X" naming), and raw sensitivity data with a matching `CELL_LINE_NAME` --
including a `"Control"` row, since sensitivity `LogFC` is computed as
fold-change relative to it:

```{r toy-paired-data}
toy_uka_full <- data.frame(
  `x.contrast` = "DrugA vs Control",
  `x.Kinase Name` = real_genes,
  `x.Median Kinase Statistic` = c(2.5, -2.0, 1.8, -1.5, 1.2, -1.0, 0.8, -0.5),
  `x.Mean Specificity Score` = 2.0,
  check.names = FALSE
)

toy_sens <- data.frame(
  CELL_LINE_NAME = rep(c("DrugA", "Control"), each = 4),
  TARGET_1 = rep(real_genes[1:4], 2),
  LN_IC50 = c(-3.5, -1.0, -3.0, -0.8, -1.0, -0.9, -1.0, -0.9)
)

toy_uka_fam <- data.frame(
  Kinase_Name = real_genes,
  Kinase_family = rep(c("FamilyX", "FamilyY"), 4)
)
```

```{r golden-score}
respath_full <- tempfile("golden_score_")
dir.create(respath_full)

result_full <- make_golden_score(
  uka = toy_uka_full,
  sens = toy_sens,
  control = "Control",
  spec_cutoff = 1.0,
  perc_cutoff = c(0.5, 0.7),
  uka_fam = toy_uka_fam,
  ppi_network = ppi_networkv12_502_kins,
  b = 50,
  respath = respath_full,
  nperms_overlap = 20,
  nperms_network = 2
)

result_full$results
```

## Reading results

`result$results` has one row per condition x `spec_cutoff` x `perc_cutoff`,
with `score_sig_network` (network-topology significance) and, for the
golden (paired) path, `score_sig_overlap`/`score_sig_overlap_fam` (kinase/
sensitivity overlap significance, individual and family-level) -- lower
means more significant.

## Output: one folder per `(spec_cutoff, perc_cutoff)` combination

`respath` above is just a base directory. Both calls go through
`networkGen::build_network_grid()`/`prepare_run_params()` internally (via
this package's own `prepare_score_run_params()`, which extends
`networkGen`'s naming rule with `nperms_network`/`relative_to`) -- so each
distinct `(spec_cutoff, perc_cutoff)` combination gets its **own** folder,
same as plain generation: two `perc_cutoff` values above means two folders,
each with `results.csv`/`metrics_permutations.csv`/`metrics_observed.csv`,
the faceted diagnostic histograms
(`Network_metrics_plots/<metric>_cutoff<cutoff>.png` -- permutation-null-
distribution per metric, faceted by condition, observed value marked with a
line), and each condition's **observed** network
(`nodes`/`edges`/`wc_df` CSVs, written by the same
`networkGen::kinograte_pg_pcsf()` mechanism `write = TRUE` uses directly).
Permutation networks (there are `nperms_network` of them per condition) are
never written -- only the observed build:

```{r list-output}
list.files(respath_kinase, recursive = TRUE)
```

For the paired (golden) path, the folder additionally nests under a
`uka`-named subfolder (since output there depends on both `uka` and `sens`):

```{r list-output-paired}
list.files(respath_full, recursive = TRUE)
```

## Safety guard for large grids

Like `networkGen::run_network_grid()`, both functions refuse to proceed
(without building anything) if the grid -- with permutations included --
would exceed `max_tasks` (default 500) networks:

```{r max-tasks, error = TRUE}
make_golden_score_kinase(
  uka = toy_uka, spec_cutoff = 1.0, perc_cutoff = c(0.5, 0.7),
  ppi_network = ppi_networkv12_502_kins, b = 50, respath = respath_kinase,
  nperms_network = 50, max_tasks = 10
)
```

## Under the hood: full flattening

Every network needed -- one observed plus `nperms_network` permuted, for
every condition x `spec_cutoff` x `perc_cutoff` combination -- is built in
one flattened batch. Set a `future` plan before calling to get actual
parallelism:

```{r batch-plan, eval = FALSE}
future::plan(future::multisession, workers = 4)
```

Without a plan set, `networkGen::generate_networks_batch()` (which both
`make_golden_score_full()`/`make_golden_score_kinase()` call internally, via
`score_conditions()`) runs sequentially -- still correct, just not parallel.
