---
title: "Getting started with networkScore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with networkScore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`networkScore` computes a permutation-test significance score for a
condition's PCSF network -- either kinase-only, or paired with sensitivity
data ("golden score"). Every network build goes through `networkGen`, so
this vignette's chunks aren't run here (they need the real, compiled `PCSF`
package -- see `networkGen`'s README).

## Kinase-only score

```{r}
library(networkScore)

# toy_uka: Tercen-style export with columns like "path.Sample", "path.Kinase Name",
# "path.Median Kinase Statistic", "path.Mean Specificity Score"
result <- make_golden_score(
  uka = toy_uka,
  spec_cutoff = 1.0,
  perc_cutoffs = c(0.5, 0.7),
  ppi_network = toy_ppi,
  b = 1.5,
  respath = tempfile("kinase_score_"),
  nperms_network = 20 # small for a quick example; use 50+ for real analyses
)

result$results
```

## Golden score (paired kinase + sensitivity)

```{r}
result_full <- make_golden_score(
  uka = toy_uka,
  sens = toy_sens,
  control = "Control",
  spec_cutoff = 1.0,
  perc_cutoffs = c(0.5, 0.7),
  uka_fam = toy_uka_fam,
  ppi_network = toy_ppi,
  b = 1.5,
  respath = tempfile("golden_score_"),
  nperms_overlap = 100,
  nperms_network = 20
)

result_full$results
```

## Under the hood: full flattening

Both calls above build, in one parallel batch, every network needed for
every condition x percentile-cutoff combination: one observed network plus
`nperms_network` permuted networks per condition. Set a `future` plan before
calling to get actual parallelism:

```{r}
future::plan(future::multisession, workers = 4)
```

Without a plan set, `networkGen::generate_networks_batch()` (which both
`make_golden_score_full()`/`make_golden_score_kinase()` call internally)
runs sequentially -- still correct, just not parallel.

## Reading results

`results$results` has one row per condition x perc_cutoff, with
`score_sig_network` (network-topology significance) and, for the golden
(paired) path, `score_sig_overlap`/`score_sig_overlap_fam` (kinase/
sensitivity overlap significance) -- lower means more significant. The run
also writes diagnostic histograms (one per metric x cutoff, faceted by
condition -- see [plot_network_metric_hist_facet()]) to
`<respath>/Network_metrics_plots/`.
