---
title: "Getting started with networkScore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with networkScore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`networkScore` computes a permutation-test significance score for a
condition's PCSF network -- either kinase-only, or paired with sensitivity
data ("golden score"). Every network build goes through `networkGen` (see
`vignette("networkGen")` for how a network is built and its parameters).
`spec_cutoff`, `perc_cutoff`, `b`, `w`, `rank_uka_abs`, and `ppi_network`
can all be tuned as vectors here -- every combination gets scored, not just
the ones a particular call happens to compare. Additional arguments (e.g.
`n`, the number of PCSF randomized runs) apply uniformly to every build,
not gridded.

```{r setup}
library(networkGen)
library(networkScore)
```

## A tiny real example

Raw UKA (kinase-activity) data, Tercen-style dotted column names, two
conditions, 8 real genes sampled from the bundled `ppi_networkv12_502_kins`
(the smallest of the three bundled reference networks -- 502 kinases,
~107K edges, vs. 1.1M for the full `ppi_networkv12` -- good enough for a
kinase-activity network and much faster for an example). The condition
column here is `Sgroup_contrast` (the default); an export that names it
differently is handled by passing `comparison_col`:

```{r toy-uka}
set.seed(1)
real_genes <- sample(unique(c(ppi_networkv12_502_kins$head, ppi_networkv12_502_kins$tail)), 8)
real_genes

toy_uka <- data.frame(
  `x.Sgroup_contrast` = rep(c("DrugA_vs_DMSO", "DrugB_vs_DMSO"), each = 8),
  `x.Kinase Name` = rep(real_genes, 2),
  `x.Median Kinase Statistic` = c(
    c(2.5, -2.0, 1.8, -1.5, 1.2, -1.0, 0.8, -0.5),
    c(-1.8, 2.2, -1.0, 1.5, -0.8, 1.0, -0.5, 0.6)
  ),
  `x.Mean Specificity Score` = 2.0,
  check.names = FALSE
)
head(toy_uka)
```

## Kinase-only score

`nperms_network` is kept small here (2) for a fast-running example; the
default (30) or more is appropriate for a real analysis. `n = 2` likewise trims PCSF's own randomized-run count
per build (default 8) for speed. `b = c(50, 75)` and `w = 10` are far above
their defaults of `2` because `toy_uka` is 8 genes drawn *at random* from
the reference network -- an unrelated set that only knits into a network
when prizes and the root-connection cost are pushed hard (see
`vignette("networkGen")`'s "How PCSF builds a network" section). Real
kinase hits from one comparison are biologically related and usually score
fine at the defaults. `perc_cutoff` **and** `b` are both vectors here --
every combination is scored, gridded against every condition (`spec_cutoff`,
`w`, `rank_uka_abs`, `ppi_network` grid the same way, see above):

```{r kinase-score}
respath_kinase <- tempfile("kinase_score_")
dir.create(respath_kinase)

result <- make_golden_score(
  uka = toy_uka,
  spec_cutoff = 1.0,
  perc_cutoff = c(0.5, 0.7),
  ppi_network = ppi_networkv12_502_kins,
  b = c(50, 75),
  w = 10,
  respath = respath_kinase,
  nperms_network = 2,
  n = 2
)
```

(`result$results`, the main output, is shown under "Reading results" below.)

## Golden score (paired kinase + sensitivity)

The paired path needs a `contrast` column identifying treatment vs. control
("X vs control" or "control vs X" naming both work), and raw sensitivity
data with a matching `CELL_LINE_NAME` -- including a `"Control"` row, since
sensitivity `LogFC` is computed as fold-change relative to it:

```{r toy-paired-data}
toy_uka_full <- data.frame(
  `x.contrast` = "DrugA vs Control",
  `x.Kinase Name` = real_genes,
  `x.Median Kinase Statistic` = c(2.5, -2.0, 1.8, -1.5, 1.2, -1.0, 0.8, -0.5),
  `x.Mean Specificity Score` = 2.0,
  check.names = FALSE
)

toy_sens <- data.frame(
  CELL_LINE_NAME = rep(c("DrugA", "Control"), each = 4),
  TARGET_1 = rep(real_genes[1:4], 2),
  LN_IC50 = c(-3.5, -1.0, -3.0, -0.8, -1.0, -0.9, -1.0, -0.9)
)

toy_uka_fam <- data.frame(
  Kinase_Name = real_genes,
  Kinase_family = rep(c("FamilyX", "FamilyY"), 4)
)
```

```{r golden-score}
respath_full <- tempfile("golden_score_")
dir.create(respath_full)

result_full <- make_golden_score(
  uka = toy_uka_full,
  sens = toy_sens,
  control = "Control",
  spec_cutoff = 1.0,
  perc_cutoff = c(0.5, 0.7),
  uka_fam = toy_uka_fam,
  ppi_network = ppi_networkv12_502_kins,
  b = 50,
  w = 10,
  respath = respath_full,
  nperms_overlap = 20,
  nperms_network = 2,
  n = 2
)
```

(`result_full$results` is shown under "Reading results" below, alongside the kinase-only table.)

## Reading results

`result$results`/`result_full$results` -- the main output of this package --
has one row per `Comparison` (the condition/contrast being scored, e.g.
`"DrugA_vs_DMSO"`) x every grid parameter, plus the output scores. Only the
parameters this particular call actually grids show up as columns --
`spec_cutoff`/`perc_cutoff`/`b`/`w`/`rank_uka_abs`/`ppi_network` each appear
only if that dimension has more than one value in this run; a parameter
fixed at a single value is dropped, since a constant column carries no
information. Lower scores mean more significant throughout.

```{r results-table-kinase, echo = FALSE}
knitr::kable(result$results, digits = 3, caption = "Kinase-only score")
```

**Kinase-only mode:**

- **`score_sig_network`** -- measures whether the top predicted kinases form
  a more interconnected network than random kinases from the same
  condition (see "Network-topology metrics" below: `rel_med_path`).
- **`score_sig_network_inv`** -- a variant of `score_sig_network` that
  focuses on the more interconnected parts of the network (see
  `rel_med_path_inv` below).
- Columns starting with **`obs_`** are the actual observed values (as
  opposed to the permutation-derived significance scores).

```{r results-table-full, echo = FALSE}
knitr::kable(result_full$results, digits = 3, caption = "Golden score (paired)")
```

**Golden-score mode** (adds sensitivity/target information on top of the
kinase-only scores above):

- **`score_sig_network`** -- measures whether the top predicted kinases form
  a more interconnected network *with the sensitivity targets* than random
  kinases from the same condition.
- **Overlap scores** (`obs_overlap`/`score_sig_overlap`) -- the overlap
  between predicted kinases and sensitivity targets, divided by the number
  of targets. The `_fam` versions (`obs_overlap_fam`/`score_sig_overlap_fam`)
  are the same comparison at the kinase-family level.

### Network-topology metrics behind `score_sig_network`

Every build -- the observed network and every permuted one -- gets four
topology statistics, written to `metrics_observed.csv` (one row per
condition x metric, from the observed build) and
`metrics_permutations.csv` (one row per condition x metric x permutation):

- **`rel_med_path`** -- median shortest-path distance between all node
  pairs, relative to network size. Smaller means a more compact network.
- **`rel_med_path_inv`** -- the same idea, but on an inverse-distance
  transform of the path lengths first, so close pairs count for more than
  far-apart ones.
- **`density`** -- fraction of all possible edges actually present.
- **`clustering`** -- mean local clustering coefficient: how much a node's
  neighbors also connect to each other (triangle-forming), averaged across
  the network.

`score_sig_network` is a real one-sided permutation p-value on
`rel_med_path`: the fraction of permutations whose `rel_med_path` is at
least as small (compact) as the observed network's, floored at
`1/nperms_network` so it's never exactly 0. `score_sig_network_inv` is the
same test on `rel_med_path_inv`. `density`/`clustering` are recorded for
inspection but don't drive either significance score.

Each `(metric, perc_cutoff)` combination also gets a diagnostic plot: a
histogram of that metric's permutation values, faceted by condition, with
the observed value marked by a red dashed line -- a quick visual check of
whether the observed network is unusually compact, rather than reading the
summary number alone. Here's `rel_med_path` at `perc_cutoff = 0.5` from the
kinase-only run above:

```{r metric-plot, echo = FALSE, results = "asis"}
rel_med_path_png <- list.files(
  respath_kinase, pattern = "^rel_med_path_cutoff0\\.5\\.png$",
  recursive = TRUE, full.names = TRUE
)[1]
# Embedded as a data URI (not knitr::include_graphics()) -- the PNG lives
# under a tempfile() path outside the vignette's own directory tree, and
# include_graphics()'s relative-path handling breaks under R CMD check's
# more deeply nested vignette-rebuild working directory, even though it
# works in a plain rmarkdown::render(). A data URI has no path to resolve.
cat(sprintf(
  '<img src="%s" style="max-width: 100%%;" alt="rel_med_path permutation-null histogram, faceted by condition">',
  base64enc::dataURI(file = rel_med_path_png, mime = "image/png")
))
```

## Output: one folder per grid combination

`respath` above is just a base directory. Each distinct `(spec_cutoff,
perc_cutoff, b, w, rank_uka_abs, ppi_network)` combination gets its **own**
folder, same as plain generation with `networkGen`: the kinase example
above grids 2 `perc_cutoff` values x 2 `b` values, so 4 folders, each with
`results.csv`/`metrics_permutations.csv`/`metrics_observed.csv`, the
faceted diagnostic histograms
(`Network_metrics_plots/<metric>_cutoff<cutoff>.png` -- permutation-null-
distribution per metric, faceted by condition, observed value marked with a
line), and each condition's **observed** network (`nodes`/`edges`/`wc_df`
CSVs). Permutation networks (there are `nperms_network` of them per
condition) are never written -- only the observed build:

```{r list-output}
kinase_files <- list.files(respath_kinase, recursive = TRUE)
length(kinase_files)
head(kinase_files, n = 2)
```

For the paired (golden) path, the folder additionally nests under a
`uka`-named subfolder (since output there depends on both `uka` and `sens`):

```{r list-output-paired}
full_files <- list.files(respath_full, recursive = TRUE)
length(full_files)
head(full_files, n = 2)
```

## Comparing scores across separate runs

`plot_score_comparison()` reads the `results.csv` under each of several
result folders and plots `score_sig_network`/`score_sig_network_inv` as
bars per `Comparison`, one facet per folder -- for comparing the *same*
comparisons across *different* parameter combinations, or entirely separate
runs, one folder per combination (matching how output is laid out
throughout this package -- see "Output: one folder per grid combination"
above). Give each folder an alias; that alias becomes its facet title, and
facets appear in the order given, not alphabetically.

Two single-combination runs, differing only in `b`, to compare against each
other (`respath_kinase` above isn't a fit here -- it's the *base* folder for
a whole grid of 4 combinations, not one combination itself):

```{r comparison-runs}
respath_b50 <- tempfile("kinase_score_b50_")
dir.create(respath_b50)
result_b50 <- make_golden_score(
  uka = toy_uka, spec_cutoff = 1.0, perc_cutoff = 0.6,
  ppi_network = ppi_networkv12_502_kins, b = 50, w = 10,
  respath = respath_b50, nperms_network = 2, n = 2
)

respath_b75 <- tempfile("kinase_score_b75_")
dir.create(respath_b75)
result_b75 <- make_golden_score(
  uka = toy_uka, spec_cutoff = 1.0, perc_cutoff = 0.6,
  ppi_network = ppi_networkv12_502_kins, b = 75, w = 10,
  respath = respath_b75, nperms_network = 2, n = 2
)
```

```{r comparison-plot, fig.width = 8, fig.height = 5.5, out.width = "100%"}
plot_score_comparison(c("b = 50" = respath_b50, "b = 75" = respath_b75))
```

Only `Comparison` values present in **every** listed folder are kept -- if
you accidentally compare folders from unrelated runs with no comparisons in
common, this refuses with a clear error instead of silently producing a
misleading empty plot.

## Safety guard for large grids

Like `networkGen::run_network_grid()`, both functions refuse to proceed
(without building anything) if the grid -- with permutations included --
would exceed `max_tasks` (default 500) networks:

```{r max-tasks, error = TRUE}
make_golden_score_kinase(
  uka = toy_uka, spec_cutoff = 1.0, perc_cutoff = c(0.5, 0.7),
  ppi_network = ppi_networkv12_502_kins, b = 50, respath = respath_kinase,
  nperms_network = 50, max_tasks = 10
)
```

## Parallel vs. sequential execution

Every network needed -- one observed plus `nperms_network` permuted, for
every condition x `spec_cutoff` x `perc_cutoff` x `b` x `w` x `rank_uka_abs` x
`ppi_network` combination -- is submitted together, so a `future` plan set
before calling parallelizes across all of it (every condition and every
permutation at once, not just across conditions):

```{r batch-plan, eval = FALSE}
future::plan(future::multisession, workers = 4)
```

Without a plan set, both functions run sequentially -- still correct, just
not parallel.
